---
title: "Lasquite_logistic_reg.rmd"
author: "Hannah Lasquite"
date: "2026-07-11"
output: html_document
---

# Fitting multiple logistic regression

*1. In RStudio, create an R Markdown file. Name it “Your Surname_logistic_reg.rmd.”*
*2. Read in the species data and remove unnecessary rows with NAs.*
```{r A12, warning = FALSE}
library(maps)
library(osmdata)
library(sf)
library(terra)
library(ggplot2)
library(tidyverse)
library(dplyr)
library(ggnewscale)

mousedeer_data <- read.csv("ph_mouse_deer.csv")
if (any(is.na(mousedeer_data)) == T){print(sum(is.na(mousedeer_data)))}
```
*There are no NAs so lets go*

*3.Generate a map showing the species' observation localities using the longitude and latitude coordinates. Check Activity 10 and use the provided codes as your guide for creating the map. Export the map as a PDF file.*
```{r A3, warning = FALSE, fig.align='center', fig.width=7, fig.height=7}
PHshp <- st_read("C:/Users/Acer/Desktop/LifeDat Projects/Act 12/PHL_adm1.shp")
Palawanshp <- PHshp %>% filter(NAME_1 %in% c("Palawan"))
PHtif <- rast("C:/Users/Acer/Desktop/LifeDat Projects/Act 12/pHMap.tif")

print(st_crs(Palawanshp) != st_crs(PHtif))

Palawanshp <- st_transform(Palawanshp, st_crs(PHtif))

Fmap <- crop(PHtif, Palawanshp) |> mask (Palawanshp)

 plot(Fmap)
 #Nice now its palawan
 
 names(Fmap) <- "elevation"
 Fmap_df <- as.data.frame(Fmap, xy = TRUE)
 

mouse_mapdata <- select(mousedeer_data, Longitude, Latitude)
mouse_mapdata

ggplot() +
    geom_tile(data = Fmap_df, aes(x,y, fill = elevation)) +
    scale_fill_gradientn(
        colors = c("#2b83ba", "#abdda4", "#ffffbf", "#fdae61", "#d7191c"),
        name = "Elevation (m)") +
    new_scale_fill() +
    geom_point(data = mouse_mapdata,
               aes(x = Longitude, y = Latitude),
               shape = 21,
               fill = "grey",
               size = 2,
               stroke = 0.5) +
    coord_equal() +
    labs(title = "Species Occurrence Area",
         x = "Longitude",
         y = "Latitude") +
   coord_sf(
    xlim = c(116.5, 120.5),
    ylim = c(7.5, 12.5),
    expand = FALSE) +
    theme_bw()
```

*4. Using the “glm()” function, perform multiple logistic regression. Ensure that the habitat types (i.e., HabitatCode; CVT - Cultivation; ESG - Early secondary growth forest; ASG - Advanced secondary growth forest; OGF - Old-growth forest) are treated as categorical variables by converting them to factors.*
```{r A4, warning = FALSE}
mouse_lm <- select(mousedeer_data, AverageDistance, Effort, HabitatCode, Detected)
mouse_lm

mouse_lm$HabitatCode <- as.factor(mouse_lm$HabitatCode)

x1 <- mouse_lm$AverageDistance
x2 <- mouse_lm$Effort
x3 <- mouse_lm$HabitatCode
y <- mouse_lm$Detected

mouse_lmdf <- data.frame(x1,x2,x3,y)
model <- glm(y~ x1+ x2 + x3,
             data = mouse_lmdf,
             family = binomial)

mouse_lmdf$Probability <- predict(model, type = "response")
mouse_lmdf$Prediction <- ifelse(mouse_lmdf$Probability >= 0.5, 1, 0)

mouse_lmdf
```

*5. After fitting the model, create a logistic regression curve plot for each independent variable used. Export the plots as PDF files.*


*Average Distance vs. Probability of Detection*
```{r A5, warning = FALSE, fig.width= 7,fig.height=7}
finalx <- data.frame(
  x1 = seq(min(mouse_lmdf$x1), max(mouse_lmdf$x1), length.out = 19),
  x2 = mean(mouse_lmdf$x2),
  x3 = levels(mouse_lmdf$x3)[1])

finalx$Probability <- predict(model, newdata = finalx, type = "response")

plot(mouse_lmdf$x1, mouse_lmdf$y,
     pch = 19,
     xlab = "Average Distance",
     ylab = "Probability",
     ylim = c(0,1))

lines(finalx$x1,
      finalx$Probability,
      col = "blue",
      lwd = 3)
```

*Effort vs. Probability of Detection*
```{r A51, warning = FALSE, fig.width= 7,fig.height=7}
finalx2 <- data.frame(
  Bx1 = mean(mouse_lmdf$x1),
  Bx2 = seq(min(mouse_lmdf$x2), max(mouse_lmdf$x2), length.out = 19),
  Bx3 = levels(mouse_lmdf$x3)[1])

finalx2$Prob <- predict(model, newdata = finalx2, type = "response")

plot(mouse_lmdf$x2, mouse_lmdf$y,
     pch = 19,
     xlab = "Effort",
     ylab = "Probability",
     ylim = c(0,1))
lines(finalx2$Bx2,
      finalx2$Prob,
      col = "blue",
      lwd = 3)
```

*Habitat Type vs. Probability of Detection*
```{r A53, warning = FALSE, fig.width= 7,fig.height=7}
finalx3 <- expand.grid(
  Cx1 = mean(mouse_lmdf$x1),
  Cx2 = mean(mouse_lmdf$x2),
  Cx3 = mouse_lmdf$x3)

finalx3$Probs <- predict(model, newdata = finalx3, type = "response")

plot(mouse_lmdf$x3, mouse_lmdf$y,
     pch = 19,
     xlab = "Habitat Type",
     ylab = "Probability",
     ylim = c(0,1))
lines(finalx2$Cx3,
      finalx2$Probs,
      col = "blue",
      lwd = 3)
```

```{r A6, warning = FALSE, fig.width= 7,fig.height=7}
summary(model)
```
