---
title: "Lasquite_randforest_reg"
author: "Hannah Lasquite"
date: "2026-07-26"
output: html_document
---
*1. In RStudio, create an R Markdown file. Name it “Your Surname_randforest_reg.rmd.”*


*2.Read in the acoustic data and remove unnecessary columns (i.e., column 1). Remember that your response variable is the species column.*


```{r A12, warning = FALSE}
library(randomForest)
library(ranger)
library(rpart)
library(rpart.plot)

setwd("C:/Users/Acer/Desktop/LifeDat Projects/Act13")

acousdat <- read.csv("call_data_12October2025.csv")
if (any(is.na(acousdat)) == T){print(sum(is.na(acousdat)))}
#No NAs lets ride

acousdat <- acousdat[, -1]
acousdat$Species <- as.factor(acousdat$Species)
```


*3. Design your Cartesian grid and perform a random forest that iterates on each hyperparameter combination.*


```{r A3, warning = FALSE}
set.seed(123)

mtry <- floor(sqrt(ncol(acousdat[,-1])))
ntree <- c(100, 200, 300, 400, 500)
min.node.size <- c(1,2,3,4,5)

rfgrid <- expand.grid(
  mtry = mtry,
  ntree = ntree,
  min.node.size = min.node.size
)

overall <- data.frame(
  mtry = numeric(),
  num.tree = numeric(),
  min.node.size = numeric(),
  accuracy = numeric(),
  oob_error = numeric()
)

for(i in 1:nrow(rfgrid)){
  rfmodel <- ranger(
    Species ~ .,
    data = acousdat,
    ntree = rfgrid$ntree[i],
    mtry = rfgrid$mtry[i],
    min.node.size = rfgrid$min.node.size[i]
  )
  
  prediction <- predict(rfmodel, data = acousdat)$predictions
  accu <- mean(prediction ==acousdat$Species)
  
  overall <- rbind(
    overall,
    cbind(rfgrid[i, ],
          accuracy = accu,
          obb_error = rfmodel$prediction.error))}


```


*4. Pick the best hyperparameter combination that minimizes the out-of-bag error.*


```{r A4, warning = FALSE}
overall[which.min(overall$obb_error), ]
```

*5. Using the optimal hyperparameter, provide a plot of variable importance and the predicted class probabilities.*


```{r A5, warning = FALSE}
set.seed(123)
Finalmodel <- randomForest(
  Species ~ .,          
  data = acousdat,
  ntree = 100,
  mtry = 3,
  importance = TRUE
)

randomForest::importance(Finalmodel)
```
```{r A51, warning = FALSE}
varImpPlot(Finalmodel)
```

```{r A52, warning = FALSE}
rpart.plot(rpart(Species ~ ., data = acousdat))

```

