---
title: "Lasquite_hclust"
author: "Hannah Lasquite"
date: "2026-06-10"
output: html_document
---
# Activity 7: K-means Clustering

## A. Hierarchical clustering

*1. In RStudio, create an R Markdown file. Name it “Your Surname_hclust.rmd.”*
```{r A1, echo = TRUE}
setwd("C:/Users/Acer/Desktop/LifeDat Projects")
library(cluster)
##This file itself is the first step.
```

*2. Read in the morphology data as a dataframe.*
```{r A2, echo = TRUE}
rawdat <- read.csv("morphology_data.csv")
```

*3. Remove unnecessary columns (e.g., sex, pop).*
```{r A3, echo = TRUE}
cleandat <- rawdat[, -c(1,2)]
scaledat <- scale(cleandat)
```

*4. Ensure that the dataframe has no NAs.*
```{r A4, echo = TRUE}
if (any(is.na(scaledat)) == T){print(sum(is.na(scaledat)))}
#If nothing is printed then we are good to go
```

**Full Academic Honesty: Stole the codes from 1-4 from my previous submission on K-Means**

*5. Before performing hierarchical clustering, calculate the “Euclidean distance” and use it as your input.*
```{r A5, echo = TRUE}
euc <- dist(scaledat, method = "euclidean")
```

*6. Perform hierarchical clustering using Ward’s and Single Linkage methods.*
```{r A6, echo = TRUE}
rownames(scaledat) <- rawdat$Pop
#For the tip labels

wardmorph <- hclust(euc, method = "ward.D2")

plot(wardmorph,
     main = "Ward's Method",
     labels = rownames(scaledat),
     cex = 0.3)

singmorph <- hclust(euc, method = "single")
plot(singmorph,
     main = "Single Linkage",
     labels = rownames(scaledat),
     cex = 0.3)

```

*7. Create a for loop function to explore Ks from 2 to 10 for each method using the Silhouette score and display the results with a plot.*
```{r A7, echo = TRUE}
wardmorph_avg <- numeric()

for (i in 2:10){
  wardmorph_clust <- cutree(wardmorph, i)
  wardmorph_sil <- silhouette(wardmorph_clust, euc)
  wardmorph_avg[i] <- mean(wardmorph_sil[, 3])}

plot(
  2:10,
  wardmorph_avg[2:10],
  xlab = "# of Clusters",
  ylab = "Silhoutte Score",
  main = "Silhoutte Score for Single Linkage",
  lwd = 2,
  cex = 1.5,
  pch = 19,
  type = "b",
)


singmorph_avg <- numeric()

for (i in 2:10){
  singmorph_clust <- cutree(singmorph, i)
  singmorph_sil <- silhouette(singmorph_clust, euc)
  singmorph_avg[i] <- mean(singmorph_sil[, 3])}

plot(
  2:10,
  singmorph_avg[2:10],
  xlab = "# of Clusters",
  ylab = "Silhoutte Score",
  main = "Silhoutte Score for Single Linkage",
  lwd = 2,
  cex = 1.5,
  pch = 19,
  type = "b",
)
```

*8. Now, generate a dendrogram for each method, highlighting the optimal K with colored boxes*
```{r A8, echo = TRUE}
plot(wardmorph,
     main = "Ward's Method",
     labels = rownames(scaledat),
     cex = 0.3)

rect.hclust(wardmorph, k = 2,
            border = c("red", "green"))

plot(singmorph,
     main = "Single Linkage",
     labels = rownames(scaledat),
     cex = 0.3)
rect.hclust(singmorph, k = 2,
            border = c("red", "green"))
```

## B. K-means vs. Hierarchical
*1. Revisit your K-means outputs from the previous activity and compare how the clusters were determined.* 
```{r B1, echo = TRUE}
#Silhouette seems more practical for me. Because the basis for the K (number of clusters), is the widest silhouette width is better than just estimating which point which has the highest drop from the elbow graph
```

*2. Are the clusters consistent? If not, what do you think causes this, and how might it influence your interpretation of biological patterns in the data?*
```{r B2, echo = TRUE}
#From both the ward's and the single linkage both produced two clusters. However the single linkage graph did have a significantly larger and significantly smaller one. 
```

