---
title: "Lasquite_hclusr"
author: "Hannah Lasquite"
date: "2026-06-17"
output: html_document
---

#A. Principal Component Analysis
*1.In RStudio, create an R Markdown file. Name it “Your Surname_hclust.rmd.”*
```{r A1, warning = FALSE}
setwd("C:/Users/Acer/Desktop/LifeDat Projects")
library(PCAtest)
library(MASS)
##This file is literally A1
```

*2. Read in the frogs’ data as a dataframe.*
```{r A2, warning = FALSE}
rawdat <- read.csv("frogdata.csv")
##Renamed for easier access
```

*3. Subset your data to retain only the “Species” column and morphological features (e.g., SVL, HL, SNL, etc.) from columns 10 through the last one (i.e., 10:X).*
```{r A3, warning = FALSE}
cleandat <-rawdat[,c(1,10:34)]
##Renamed for easier access
```

*4. Check the class and structure of the subset data. How many morphological features are present in the dataset?*
```{r A4, warning = FALSE}
str(cleandat)
class(cleandat)
##Congrats, its a dataframe

print(ncol(cleandat) - 1)
##Minus 1 for the species column, there are 25 morphological features in the data set
```

*5.Ensure the dataframe has no NAs.*
```{r A5, warning = FALSE}
if (any(is.na(cleandat)) == T){print(sum(is.na(cleandat)))}
##Checks for NAs and prints the amount if any. If nothings printed its good
```

*6.Scale the data using only the numeric columns, then compute the pairwise correlations. Do not further subset the data.*
```{r A6, warning = FALSE}
scaledat <- scale(cleandat[, 2:26])
pairdat<- cor(scaledat)

round(pairdat, 2)
##For easier reading
```

*7. Use the correlation matrix to calculate the eigenvectors and the proportion of variance (expressed in percentage). How many principal components were produced? What is the proportion of variance explained by components 1–4?*
```{r A7, warning = FALSE}
eigdat <- eigen(pairdat)
propvardat <- eigdat$values / sum(eigdat$values)

percentdat <- (eigdat$values / sum(eigdat$values)) * 100
##The percentage version of propvardat
##There were 25 PCs produced

eigvectdat <- eigdat$vectors
scoredat <- scaledat %*% eigvectdat
head(scoredat)

sum(percentdat[1:4])

##The amount of proportion explained by PCs 1-4
```
*8. Now, create a 1x2 plot showing the screeplot and the PCA biplot. You can use either base R or ggplot2. You can be creative here by coloring the point shapes per species and adding a legend. Consider the high data-ink ratio you learned from the lecture.*
```{r A8, warning = FALSE}
par(mfrow = c(1, 2))
barplot(propvardat,
        names.arg = paste0("PC", 1:25),
        ylim = c(0, 1),
        col = "darkgreen",
        main = "Explained Variance (Scree Plot)",
        ylab = "Proportion of Variance")


colordat <-c(P_guiting = "green", P_pygmaeus = "red")
colordat2 <- as.factor(colordat)

plot(scoredat[,1], scoredat[,2],
     xlab = "PC1",
     ylab = "PC2",
     main = "PCA Score Plot (PC1 vs PC2)",
     pch = 17,
     col = colordat)
legend("topright",
       legend = c("P_guiting","P_pygmaeus"),
       col = colordat,
       pch = 17)
dev.off()
```

#B. Broken-stick vs PCAtest
*1. Examine the variance explained by principal components using the Broken-stick method and the PCAtest package. What is the difference between these two methods for assessing the importance of principal components?*
```{r B1, warning = FALSE}
bs1 <- function(p){
  sapply(1:p, function(k){
    sum(1 / (k:p)) / p })}

bs2 <- bs1(ncol(scaledat))

compardat <- data.frame(
  PC = paste0("PC", 1:25),
  Observed = propvardat,
  BrokenStick = bs2
)

compardat

plot(propvardat,
     type = "b",
     pch = 19,
     ylim = c(0, max(propvardat)),
     xlab = "Principal Component",
     ylab = "Proportion of Variance",
     main = "Broken-Stick Analysis")

lines(bs2,
      type = "b",
      pch = 17,
      col = "red")

legend("topright",
       legend = c("Observed", "Broken Stick"),
       pch = c(19, 17),
       col = c("black", "red"))

pcaresults <- PCAtest(scaledat, 100, 100, 0.5, varcorr = FALSE, counter = FALSE, plot = TRUE)
```

*2.Create a barplot showing the results of Broken-stick analysis and perform a significance test for principal components using PCAtest, following the instructions at https://arleyc.github.io/PCAtest/.*
```{r B2, warning = FALSE}
barplot(bs2,
        col = "steelblue",
        main = "Barplot of Broken Stick Analysis")
```