---
title: "Lasquite_K-means"
author: "Hannah Lasquite"
date: "2026-06-08"
output: html_document
---

# Activity 7: K-means Clustering

## A. K-means (single K)

*1.In RStudio, create an R Markdown file. Name it “Your Surname_K-means.rmd.”*

```{r A1, echo = TRUE}
setwd("C:/Users/Acer/Desktop/LifeDat Projects")
##This file itself is the first step.
```

*2. Read in the morphology data as a dataframe.*
```{r A2, echo = TRUE}
rawdat <- read.csv("morphology_data.csv")
```

*3.Remove the sex and population columns.*
```{r A3, echo = TRUE}
cleandat <- rawdat[, -c(1,2)]
```

*4. Ensure that the dataframe has no NAs.*
```{r A4, echo = TRUE}
if (any(is.na(cleandat)) == T){print(sum(is.na(cleandat)))}
#If nothing is printed then we are good to go
```

*5. Prior to performing K-means, standardize all values using the “scale” function.*
```{r A5, echo = TRUE}
scaledat <- scale(cleandat)
set.seed(63)

```

*6.Perform K-means with 3 cluster centers using 25 initial random centroids*
```{r A6, echo = TRUE}
kmeandat <- kmeans(scaledat, centers = 3, nstart = 25)
```

*7. Show the cluster assignment and a summary of results by simply printing them to the console.*
```{r A7, echo = TRUE}
kmeandat
```

*8. Create a scatter plot with points colored by the assigned cluster.*
```{r A8, echo = TRUE}
x <- scaledat[ ,1]
y <- scaledat[ ,2]
plot(
   x, y,
   col = kmeandat$cluster,
   pch = 20,
   xlab = "TD",
   ylab = "SVL",
   main = "K-means Clustering of Morphology Data"
 )
```

## B. K-means (loop)

*1. Now, create a loop function to perform K-means with cluster centers from 2 to 10.*
```{r B1, echo = TRUE}
loopK <- 1:10
elbloopK <- numeric(length(loopK))

for (i in loopK){
  multK <- kmeans(scaledat, center = i, nstart = 25)
  elbloopK[i] <- multK$tot.withinss
}


```

*2. From the output of your loop function, create an elbow curve plot to determine the optimal K.*
```{r B2, echo = TRUE}
plot(
  loopK,
  elbloopK,
  xlab = "# of Clusters",
  ylab = "Total Sum of Squares",
  main = "Morphology Elbow Plot",
  lwd = 2,
  cex = 1.5,
  pch = 19,
  type = "b"
)

##The bend starts to become shallow at 2 clusters
```

*3. Recreate steps 6–8 above using the optimal K determined by the elbow curve.*
```{r B3, echo = TRUE}
#Step 6
kmeandat2 <- kmeans(scaledat, centers = 2, nstart = 25)

#Step 7
kmeandat2

#Step 8
x2 <- scaledat[ ,1]
y2 <- scaledat[ ,2]
plot(
   x2, y2,
   col = kmeandat2$cluster,
   pch = 20,
   xlab = "TD",
   ylab = "SVL",
   main = "K-means Clustering of Morphology Data"
 )
```