---
title: "Lasquite_GIS"
author: "Hannah Lasquite"
date: "2026-06-23"
output: html_document
---
#A. Principal Component Analysis
*1. In RStudio, create an R Markdown file. Name it “Your Surname_gis.rmd.”*
```{r A1, warning = FALSE}
library(maps)
library(osmdata)
library(sf)
library(terra)
library(ggplot2)
library(tidyverse)
library(dplyr)
library(ggnewscale)
setwd("C:/Users/Acer/Desktop/LifeDat Projects/Act 9")
```

*2. Go to the Global Biodiversity Information Facility (GBIF) website. Create your account and download your assigned species records below as a CSV file (see Table). Use the species name when you search the data.*
```{r A2, warning = FALSE}
rawdata <- read.csv("Occurance Data.csv")
Mapdata <- select(rawdata, decimalLongitude, decimalLatitude, stateProvince)
Mapdata
```

*3. Clean the data by removing records found outside the reported species range.*
```{r A3, warning = FALSE}
#There's a ton of NA data with no coordinates or no Province. If there is no province I searched the coordinates up to check which Province it fell in. If there was no coordinates I removed it completely.

CleanMap <- Mapdata %>% drop_na(decimalLongitude, decimalLatitude)
#Removes any rows with missing Longitude or Latitude values


#Row 7 says the Province is Davao, Davao is no longer a singular province. Coordinates is in Davao Oriental.
CleanMap[7, "stateProvince"] <- "Davao Oriental"

#Rows 4, 5, 9, 11, 12 have empty Provinces. I am filling them in manually
CleanMap[4:5, "stateProvince"] <- "Agusan del Sur"
CleanMap[4:5, "stateProvince"] <- "Agusan del Sur"
CleanMap[9, "stateProvince"] <- "Misamis Oriental"
CleanMap[11:12, "stateProvince"] <- "Zamboanga del Sur"
CleanMap

#First two rows say the State Province is Autonomous Region in Muslim Mindanao, it does not correlate with its coordinates. The coordinates are actually in Mandaluyong?? No clue why but regardless its out of range so I sliced it away
FCleanMap <- CleanMap[-c(1:2), ]
FCleanMap
  
```

*4. Go to the DIVA-GIS website and download the Philippine administrative boundaries/areas. The zip file contains multiple shapefiles (adm0–3), representing different boundary levels (0-country; 1-provincial; 2-municipal; 3-barangay).*
*5. Get only the provincial level, and subset the provinces that contain only distribution records for your assigned species.*
```{r A45, warning = FALSE}
neededProvince <- FCleanMap[, "stateProvince"]

provdata <- st_read("PHL_adm1.shp")
provcol <- select(provdata, NAME_1)

finalprov <- provcol %>% 
  filter(NAME_1 %in% neededProvince)
finalprov
```

*6.Go back to the DIVA-GIS website. Download the Philippine digital elevation model (“Elevation [country mask]”), then clip it using the provinces you selected in step 5.*
```{r A6, warning = FALSE}
PHMap <- rast("phMap.tif")

print(st_crs(finalprov) != st_crs(PHMap))
#Just to check if the coordinate systems are the same

FinalMap <- crop(PHMap, finalprov) |> mask(finalprov)
```

#B. Create your first map
*1. Now, put all the data together into a map, following the pre-lab lecture. Make sure that data layers (points, polygons, DEMs) are arranged properly. Be creative!”*
```{r B1, warning = FALSE}
MapDF <- as.data.frame(FinalMap, xy = TRUE)

OccuranceMap <- ggplot() +
  geom_tile(data = MapDF, aes(x,y, fill = Band_1)) +
  scale_fill_gradientn(
    colors = c("#2b83ba", "#abdda4", "#ffffbf", "#fdae61", "#d7191c"),
    name = "Elevation (m)") +
new_scale_fill() +
  geom_point(data = FCleanMap,
             aes(x = decimalLongitude, y = decimalLatitude, fill = stateProvince),
             shape = 21,
             color = "black",
             size = 3,
             stroke = 1.2) +
  coord_equal() +
  labs(title = "Southern Dwarf Kingfisher Occurance",
       x = "Longitude",
       y = "Latitude",
       fill = "Province") +
  theme_bw()

ggsave("OccuranceMap.pdf", width = 8, height = 8, units = "in")
```
