---
title: "Lasquite_linear_reg.rmd"
author: "Hannah Lasquite"
date: "2026-07-10"
output: html_document
---

# A. Simple linear regression

*1. In RStudio, create an R Markdown file. Name it “Your Surname_linear_reg.rmd.”*
```{r A1, warning=FALSE}
library(terra)
library(sf)
#This file itself is step 1
```

*2. Read in the species data and remove unnecessary rows with NAs.*
```{r A2, warning=FALSE}
raw_herpdata <- read.csv("herpetozoa-033-095-s001.csv")
raw_herpdata

comp_herpdata <- complete.cases(raw_herpdata)
head(raw_herpdata[comp_herpdata, ])
herpdata <- raw_herpdata[comp_herpdata, ]
#using this method since tidyverse and terra do NOT like each other

if (any(is.na(herpdata)) == T){print(sum(is.na(herpdata)))}
#Me and my anxious butt always needs to double check
```

*3. Read in the elevation data (ph_strm_geo.tif).*
```{r A3, warning=FALSE}
PHMap <- rast("phMap.tif")
```

*4. From the species data, subset the longitude and latitude coordinates. Then convert the coordinates to a spatial vector using the “vect” function from the “terra” package.*
```{r A4, warning=FALSE}
coord <- vect(herpdata[,2:3])
coord
```

*5. Using the converted coordinates, extract elevation data for each point with the “extract” function in the “terra” package. The extracted data will have two columns: “ID” and “Band_1.” The “Band_1” column contains the elevation.*
```{r A5, warning=FALSE}
comp_coord <- extract(PHMap, coord)
```

*6. Subset the species data to retain only the species richness column (SPPTotal). Then combine this column with the extracted elevation. Rename “Band_1” to “elevation.”*
```{r A6, warning=FALSE}
specrich <- herpdata[, 6]
comb_coord_specrich <- data.frame(
  Species_Richness = specrich,
  Elevation = comp_coord[,2])
```

*7. Now, perform simple linear regression using the code from the pre-lab lecture, including a plot. What can you infer about the relationship between species and elevation?*
```{r A7, warning=FALSE}
x <- comb_coord_specrich$Elevation
y <- comb_coord_specrich$Species_Richness

xbar <- mean(x)
ybar <- mean(y)

b <- sum((x - xbar) * (y - ybar)) / sum((x - xbar)^2)
a <- ybar - b * xbar
yhat <- a + b * x

singplot_dat <- data.frame(
  Species_Richness = x,
  Elevation = y,
  Predicted = round(yhat,2)
)

plot(x, y,
     pch = 19,
     col = "blue3",
     xlab = "Elevation Level",
     ylab = "Species Richness",
     main = "Linear Regression of Species Richness and Elevation")
abline(a, b,
       col = "brown2",
       lwd = 2)
points(x, yhat,
       pch = 17,      # Triangle
       col = "darkgreen",
       cex = 1.2)
segments(x0 = x, y0 = y,
         x1 = x, y1 = yhat,
         lty = 2,
         col = "skyblue4")
legend("topright",
       legend = c("Observed", "Predicted", "Regression Line"),
       pch = c(19, 17, NA),
       lty = c(NA, NA, 1),
       col = c("blue3", "darkgreen", "brown2"),
       lwd = c(NA, NA, 2),
       bty = "n")

SSE <- sum((y - yhat)^2)
SST <- sum((y - ybar)^2)
R2 <- 1 - (SSE / SST)
R2
```
*Species Richness and Elevation Level seems to have an inverse relationship wherein species richness decreases as elevation level increases. However based on the r^2 value, only 19% of the variation in the data can be accounted for using our current model.*

# B. Multiple linear regression

*1.Read in both annual precipitation and temperature data and extract the data following the instructions from A4–5.*
```{r B1, warning=FALSE}
AnnualTemp <- rast("annual_temperature.tif")
AnnualPrecip <- rast("annual_precipitation.tif")

compTemp <- extract(AnnualTemp, coord)
compPrecip <- extract(AnnualPrecip, coord)
```

*2. Combine the extracted data with the species and elevation data into a single dataframe.*
```{r B2, warning=FALSE}
temp_precip <- data.frame(
  Temperature = compTemp[,2],
  Precipitation = compPrecip[,2]
)

MultLin_data <- cbind(comb_coord_specrich, temp_precip)
MultLin_data
```

*3. Perform multiple linear regression using the “lm” function.*
```{r B3, warning=FALSE}
MultLin <- lm(MultLin_data$Species_Richness ~ MultLin_data$Elevation + MultLin_data$Temperature + MultLin_data$Precipitation)
coef(MultLin)

a <- coef(MultLin)[1]
b <- coef(MultLin)[2]
c <- coef(MultLin)[3]
d <- coef(MultLin)[4]

cat("Regression Equation:\n")
cat("y =", round(a,3), "+", round(b,3), "x", "+", round(c,3), "x1", "+", round(d,3), "x2\n\n")

multyhat <- predict(MultLin)

residuals <- residuals(MultLin)

summary(MultLin)$r.squared

Multresult <- data.frame(
  Elevation = MultLin_data$Elevation,
  Precipitation = MultLin_data$Precipitation,
  Temperature = MultLin_data$Temperature,
  Observed = MultLin_data$Species_Richness,
  Predicted = round(multyhat,2),
  Residual = round(residuals,2)
)

Multresult

summary(MultLin)
```

*4. Which relationship is statistically significant?*

*Only Temperature and Precipitation have a statistically significant relationship because they have P-values less than 0.05. Elevation is not a significally different enough to be considered an important factor.*